# Reproducible Research: Peer Assessment 1



## Loading and preprocessing the data
```{r loading}
raw.data <- read.csv("activity.csv")
raw.data$date <- as.Date(raw.data$date)
```

## What is mean total number of steps taken per day?
```{r first crude analysis}
hist(tapply(raw.data$steps, raw.data$date, sum, na.rm = TRUE),
     main = "Steps per day",
     ylab = "frequency",
     xlab = "steps")
mean(tapply(raw.data$steps, raw.data$date, sum, na.rm = TRUE))
median(tapply(raw.data$steps, raw.data$date, sum, na.rm = TRUE))
```

## What is the average daily activity pattern?
```{r daily activity}
plot(tapply(raw.data$steps, 
            raw.data$interval, 
            mean, na.rm = TRUE),
            type = "l", 
     main = "Average daily activity pattern",
     ylab = "n steps (mean)",
     xlab = "interval")
```

interval with maximum number of steps:
```{r daily max}
average <- tapply(raw.data$steps, raw.data$interval, mean, na.rm = TRUE)
average[average == max(average)]
```


## Imputing missing values

# Total number of rows with NAs:
```{r missings}
nrow(raw.data)- sum(complete.cases(raw.data))
```

# Replacing missing values with mean per interval
```{r impute}

library("reshape2")

molten.data <- melt(raw.data,
                    id = c("interval", "date"))

mean.data <- dcast(molten.data, interval ~ variable,
                   mean, na.rm = TRUE)

impu.data <- merge(raw.data, mean.data,
              by.x = "interval", by.y = "interval")

impu.data$steps <- impu.data$steps.x
impu.data$steps[is.na(impu.data$steps)] <- impu.data$steps.y[is.na(impu.data$steps)]
colnames(impu.data)[2] <- "rawsteps"
colnames(impu.data)[4] <- "meansteps"
```


# Steps per day:
```{r daily steps imputed}
hist(tapply(impu.data$steps, raw.data$date, sum, na.rm = TRUE),
     main = "Steps per day (missings imputed)",
     ylab = "frequency",
     xlab = "steps")
```
After imputation, the number of steps per day was considerably increased, and
the distribution pattern changed to a right-skewed distribution.

# Comparing mean and median (complete case vs. imputed analysis):
```{r mean}
mean(tapply(impu.data$rawsteps, impu.data$date, sum, na.rm = TRUE))
mean(tapply(impu.data$steps, impu.data$date, sum, na.rm = TRUE))

median(tapply(impu.data$rawsteps, impu.data$date, sum, na.rm = TRUE))
median(tapply(impu.data$steps, impu.data$date, sum, na.rm = TRUE))
```
Mean and median are increased roughly by 1000 steps after imputation. 


## Are there differences in activity patterns between weekdays and weekends?

```{r weekdays}
impu.data$weekend <- weekdays(impu.data$date)=="Saturday" | 
  weekdays(impu.data$date) == "Sunday"

impu.data$weekday[impu.data$weekend] <- "weekend"
impu.data$weekday[!impu.data$weekend] <- "weekday"
```

```{r weekend vs weekday}
par(mfrow=c(2,1))
plot(tapply(raw.data$steps[impu.data$weekday == "weekday"], 
            raw.data$interval[impu.data$weekday == "weekday"], 
            mean, na.rm = TRUE),
            type = "l",
     main = "Average daily activity pattern (weekdays)",
     ylab = "n steps (mean)",
     xlab = "interval")
plot(tapply(raw.data$steps[impu.data$weekday == "weekend"], 
            raw.data$interval[impu.data$weekday == "weekend"], 
            mean, na.rm = TRUE),
            type = "l",
     main = "Average daily activity pattern (weekend)",
     ylab = "n steps (mean)",
     xlab = "interval")
```